The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
6.2.1

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas. This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.

\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive on the water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
        The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.

>>>>

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
Ultimately, such measurements will shed light on the question how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetaty climates.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition. From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}. But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation. Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2022NatAs...6.1296K}. Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}. Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution. Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal. However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of oceans \citep{2018AJ....155..195W} on even the most irradiated exoplanets \citep[see the discussion in][]{2022ApJ...938L...3L}.

\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive to the water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern: The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection. Howver, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus a straightforward trend is difficult to discern.
